---
title: "ama_workspace"
output: html_document
date: "2026-09-19"
---

---
title: "miles_rep_1"
author: "nanaama-sak"
date: "2026-09-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# MILES Study Replication

## Question: Is the positive association between butyrate-producers in

the gut microbiome and insulin homeostasis (here, insulin sensitivty)
consistent across datasets? This is replication of the analysis done by
The Microbiome and Insulin Longitudinal Evaluation Study
[<https://pmc.ncbi.nlm.nih.gov/articles/PMC9630078/>]

## Dataset: Integrated Personal Omics Profiling (iPOP)! The original

paper ran on the (unavailable) iHMP/HMP2 (integrative Human Microbiome
Project/Human Microbiome Project 2)

### Set-up

Loading libraries, phyloseq object

```{r librarysetup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install("phyloseq")
library(phyloseq)
# BiocManager::install("dplyr")
library(dplyr)
# install.packages("tidyr")
library(tidyr)
library(ggplot2)
# install.packages("here")
library(here)
library(microshades)
install.packages("tidyverse")
library(forcats)
library(tidyverse)
```

```{r load-ps}
ps_file <- here("tutorial2_objects.RData")
load(ps_file)
ps #it is loading correctly!
sample_variables(ps) #these are the metadata columns we can group by
```

### Data check
*Is our data similar enough to the MILES dataset (HMP)?*
1. Do we have a gut microbiome measurement and insulin phenotype on the same participants? 

2. do we have the same genera as MILES HMP? 

3. Do we have enough independent observations (enough participants at a given
timepoint)

```{r datacheck}
nsamples(ps) #855 samples
ntaxa(ps) #96 features/taxa
taxonomy <- as.data.frame(as(tax_table(ps), "matrix"))
table(taxonomy$Rank)

#is our data relative or raw?
range(otu_table(ps)) #we are relative
```

Yes! We have a gut microbiome measurement. Now *what about insulin?*

```{r datacheck2}
sample_variables(ps) #ir_is_classification is what its called.

#is it on the same pts?
table(sample_data(ps)$ir_is_classification)

#ir_is_classification is in sample data, so it will correspond with abundance table. 
```

Yes! Insulin phenotypes are known per pt. Note that 192 individuals have unknown classification... so will likely filter these out? Later?

*Do we have the same genera studied in MILES?*

```{r datacheck3}
#MILES looks at alistipes, anaerostipes, copracoccus, flavonifractor, odoribacter, oscillibacter, pseudoflavonifracto

hmp_genera <- taxonomy$Taxon[taxonomy$Rank == "Genus"]
hmp_genera
miles_genera <- c("Alistipes", "Anaerostipes", "Coprococcus", "Flavonifractor", "Odoribacter", "Oscillibacter", "Pseudoflavonifractor") #lookuptable?? for miles genera?
miles_genera

generacheck <- data.frame(Genus = miles_genera, in_hmp = miles_genera %in% hmp_genera)

generacheck #so we have all EXCEPT anaerostipes 
```

SO! We have 6 of the 7 relevant genera studied. I feel this is fine.
Onward we push.

*Do we have enough independent observations?*

```{r datacheck3}
observations <- as.data.frame(as(sample_data(ps), "matrix")) #trying to pull out sample data
observations

#i want to know how many visits per subject
observations |>
  group_by(VisitID) |> #96 subjects
  tally() #this is showing how many subjects came to a given visit. Visit #1 had the most ppl come. 
```

VisitID 01 had the most pts involved (n=69). If we are using a single
time point we should use this visit.

So we will eventually need to filter both by:
1. presence at visit #1
2. unknown insulin phenotype.

We will also not be able to make conclusions about genera Aeroyadayada.
But I would argue our data match enough

### Building the Analysis Subset

#### Narrow to one visit*

```{r visit1only}
visit01 <- subset_samples(ps, VisitID == "01")
nsamples(visit01) #correct # of participants (n=69)
```

#### Narrow to just IS or IR
```{r visit01-IRIS-status}
irisvisit01 <- subset_samples(visit01, ir_is_classification %in% c("IR", "IS"))

nsamples(irisvisit01) #44 people have IR/IS status known from visit 1

table(sample_data(irisvisit01)$ir_is_classification) #28 people insulin resistant, #16 people insulin sensitive

```

Our Sample Set now is 44 participants (IR or IS, visit 1). Not a great balance (IR = 28, IS = 16). Keep in mind.

#### filter to genus
```{r}
irisv1genera <- subset_taxa(irisvisit01, Rank == "Genus")
ntaxa(irisv1genera) #45 genera

as.data.frame(tax_table(irisv1genera))
```

#### sidequest filter to those w/ serum metabolome data
```{r}
pilot_ps <- subset_samples(irisv1genera, SubjectID %in% c("ZK112BX", "ZLZQMEV", "ZM7JY3G", "ZPF36E2", "ZTJ7L7Z", "ZUPAQVU", "ZVBQY1N", "ZVM4N7A", "ZW61YGW", "ZWCZHHY", "ZWFDEY0", "ZY7IW45", "K4CK8Y", "ZKFV71L", "ZLZNCLZ", "ZN3TBJM", "ZN9YTFN", "ZPMBHPS"))

# pilot_ps2 <- subset_samples(irisv1genera, SubjectID %in% c("ZN9YTFN-01", "ZN3TBJM-01", "ZM7JY3G-01", "ZLZQMEV-01", "ZLZNCLZ-01", "ZY7IW45-01", "ZWFDEY0-01", "ZWCZHHY-01", "ZW61YGW-01", "ZVM4N7A-01", "ZVBQY1N-01", "ZUPAQVU-01", "ZTJ7L7Z-01", "ZPMBHPS-01", "ZPF36E2-01", "ZKFV71L-01", "ZK4CK8Y-01", "ZK112BX-01"))

nsamples(pilot_ps) 

head(as.data.frame(tax_table(pilot_ps)))
```


```{r}
# #and now scrapping non-relevant genera
# iris_v1_miles_genera <- subset_taxa(irisv1genera, Rank %in% c("Alistipes", "Coprococcus", "Flavonifractor", "Odoribacter", "Oscillibacter", "Pseudoflavonifractor"))
```

```{r cleaning-and-vibe-check}
sum(taxa_sums(irisv1genera) == 0) #no genus with 0 samples
irisv1genera <- prune_taxa(taxa_sums(irisv1genera) > 0, irisv1genera)
ntaxa(irisv1genera)
irisv1genera

summary(sample_sums(irisv1genera)) # doesn't quite equal 1
```

### IS/IR Abundance comparison (genus level)

To determine (1) what abundance looks like in each group, and (2) how abundances compare across group

#### Normalize & melt our subsetted ps
```{r norm-and-melt}
miles_ps_rel <- transform_sample_counts(irisv1genera, function(x) x / sum(x)) #this normalizes our ps
all(abs(sample_sums(miles_ps_rel) -1) < 1e-9) #every sample is 1 = true!

#melting into a df

miles_df <- psmelt(miles_ps_rel)
dim(miles_df) #44 samples x 45 genera = 1980 rows
miles_df
```

doing the same with pilot ps
```{r}
pilot_ps_rel <- transform_sample_counts(pilot_ps, function(x) x / sum(x)) #this normalizes our ps
all(abs(sample_sums(pilot_ps_rel) -1) < 1e-9) #every sample is 1 = true!

#melting into a df

pilot_df <- psmelt(pilot_ps_rel)
dim(pilot_df) #17 samples x 45 genera = 765 rows
pilot_df

any(pilot_df$sspg == 0) #sspg is steady state glucose reading... this helps us make glucose a continuous variable
```

#### Genera profile within group

Calculating the mean relative abundace of each genus within IR and IS groups, then visualizing
```{r }
genus_means <- miles_df |>
  group_by(ir_is_classification, Taxon) |>
  summarise(mean_abund = mean(Abundance, .groups = "drop")) #made a new category called mean_abund)
genus_means
ggplot(genus_means, aes(x = ir_is_classification, y = mean_abund, fill = Taxon)) + geom_col() +
  labs(
    title = "Mean genus-level composition by Insulin Status",
    x = "IR/IS Classification",
    y = "Mean relative abundance"
  ) + 
  theme_minimal() +
  theme(legend.key.size = unit(0.3, "cm"),
        legend.text = element_text(size=6))
```


microshades attempt??
```{r}
remotes::install_github("mikemc/speedyseq")
mdf_prep <- prep_mdf(irisvisit01)
head(as.data.frame(tax_table(irisvisit01)))

```


```{r }
genus_means <- miles_df |>
  group_by(ir_is_classification, Taxon) |>
  summarise(mean_abund = mean(Abundance, .groups = "drop")) #made a new category called mean_abund)
genus_means
ggplot(genus_means, aes(x = ir_is_classification, y = mean_abund, fill = Taxon)) + geom_col() +
  labs(
    title = "Mean genus-level composition by Insulin Status",
    x = "IR/IS Classification",
    y = "Mean relative abundance"
  ) + 
  theme_minimal() +
  theme(legend.key.size = unit(0.3, "cm"),
        legend.text = element_text(size=6))
```
```

First, making ps of just the top 

```{r}
top_genera <- names(sort(taxa_sums(irisv1genera), decreasing = TRUE)[1:20])
ps_top_twenty <- prune_taxa(top_genera, irisv1genera)
top_twenty_df <- psmelt(ps_top_twenty) #melting it

#calculating means
twenty_means <- top_twenty_df |>
  group_by(ir_is_classification, Taxon) |>
  summarise(mean_abund = mean(Abundance, .groups = "drop")) #made a new category called mean_abund)


ggplot(twenty_means, aes(x = ir_is_classification, y = mean_abund, fill = Taxon)) + geom_col() +
  labs(
    title = "Top 20 Mean genus-level composition by Insulin Status",
    x = "IR/IS Classification",
    y = "Mean relative abundance"
  ) + 
  theme_minimal() +
  theme(legend.key.size = unit(0.3, "cm"),
        legend.text = element_text(size=6))


```

same for pilot
```{r}
# genus_means_pilot <- pilot_df |>
#   group_by(ir_is_classification, Taxon) |>
#   summarise(mean_abund = mean(Abundance, .groups = "drop")) #made a new category called mean_abund)
# genus_means_pilot
# 
# ggplot(genus_means_pilot, aes(x = ir_is_classification, y = mean_abund, fill = Taxon)) + geom_col() +
#   labs(
#     title = "PILOT: Mean genus-level composition in Patients with Metabolomic Data, by Insulin Status",
#     x = "IR/IS Classification",
#     y = "Mean relative abundance"
#   ) + 
#   theme_minimal() +
#   theme(legend.key.size = unit(0.3, "cm"),
#         legend.text = element_text(size=6))
# 
# pilot_df
```

```{r}
##narrow to just clostridium associated genera?

genus_means %>%
  filter(Taxon %in% c("Clostridium.IV", "Clostridium.XI", "Clostridium.XIVa", 
  "Clostridium.XIVb", "Clostridium.XVIII", "Clostridium.sensu_stricto", "unclassified_Clostridiales", "unclassified_Clostridiales_Incertae.Sedis.XIII")) %>%
  ggplot(aes(x = ir_is_classification, y = mean_abund, fill = Taxon)) + 
  geom_col(position = "dodge") +  # Side-by-side bars
  labs(
    title = "Mean Clostridium Cluster Composition by Insulin Status",
    x = "IR/IS Classification",
    y = "Mean relative abundance"
  ) + 
  theme_minimal() +
  theme(legend.key.size = unit(0.3, "cm"),
        legend.text = element_text(size=6))
```


```{r}
genus_means |>
  group_by(ir_is_classification) |>
  slice_max(mean_abund, n = 45) |> #showing top 5 genera per group
  mutate(percent = round(100* mean_abund, 1)) |>
  select(ir_is_classification, Taxon, percent)
```
Bacteroides most abundant (39.7%, followed by Faecalibacterium & Lachnospiracea) --  IN ALL pts


now for pilot
```{r}
#relative abundance numbers are__ ??
genus_means_pilot |>
  group_by(ir_is_classification) |>
  slice_max(mean_abund, n = 5) |> #showing top 10 genera per group
  mutate(percent = round(100* mean_abund, 1)) |>
  select(ir_is_classification, Taxon, percent)

mean_genera_pilot <- genus_means_pilot |>
  group_by(ir_is_classification) |>
  slice_max(mean_abund, n = 45) |> #
  mutate(percent = round(100* mean_abund, 1)) |>
  select(ir_is_classification, Taxon, percent)

mean_genera_pilot
```
```{r}
pilot_obs <- as.data.frame(as(sample_data(pilot_ps), "matrix"))
pilot_obs
```
#### Abundance indices (PILOT)
```{r}
pilot_matrix <- as(otu_table(pilot_ps), "matrix")

if(taxa_are_rows(pilot_ps)) pilot_matrix <- t(pilot_matrix)

pilot_alpha <- data.frame(
  SampleID = rownames(pilot_matrix),
  Observed = rowSums(pilot_matrix > 0),                        # richness: genera present
  Shannon  = vegan::diversity(pilot_matrix, index = "shannon"),
  Simpson  = vegan::diversity(pilot_matrix, index = "simpson"),
  Group    = factor(sample_data(pilot_ps)$ir_is_classification[
                      match(rownames(pilot_matrix), sample_names(pilot_ps))],
                    levels = c("IR", "IS"))
)

head(pilot_alpha)
```
We calculated alpha diversity metrics! note to self to amend this to pilot_df.

```{r}
# Subset Clostridium only
clos_ps <- subset_taxa(pilot_ps, Taxon %in% c(
  "Clostridium.IV", "Clostridium.XI", "Clostridium.XIVa", 
  "Clostridium.XIVb", "Clostridium.XVIII", "Clostridium.sensu_stricto", "unclassified_Clostridiales", "unclassified_Clostridiales_Incertae.Sedis.XIII" 
))

# Relative abundance within Clostridium
clos_rel <- transform_sample_counts(clos_ps, function(x) x/sum(x))
```


```{r}
# Stacked bar
plot_bar(clos_rel, x = "ir_is_classification", fill = "Taxon")
```


```{r}
# Box plot of each cluster by group
clos_melt <- psmelt(clos_rel)
ggplot(clos_melt, aes(x = ir_is_classification, y = Abundance, fill = Taxon)) +
  geom_boxplot() +
  facet_wrap(~Taxon)
```


