---
title: "Metabolomics Workbench: Querying the API from R"
author: "Keisha Harrison"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tutorial overview

The [Metabolomics Workbench](https://www.metabolomicsworkbench.org/) is a public NIH-funded repository of metabolomics data and tools. The `metabolomicsWorkbenchR` package gives you a programmatic R interface to its REST API, so you can pull data and metadata directly into your R session instead of clicking through the website.

By the end of this tutorial you will be able to:

- Install `metabolomicsWorkbenchR` from Bioconductor
- Inspect the catalog of valid query types (`output_item`)
- Run a name-based RefMet query for a single analyte

**Prerequisites:** R 4.0+ and RStudio. Internet access (the API is queried over HTTPS).

---

# Step 1: Set up the working environment

## 1.1 Install the package

`metabolomicsWorkbenchR` lives on **Bioconductor**, not CRAN, so we install it through `BiocManager` rather than `install.packages()`.

The `if (!require(...))` pattern below means: *"if BiocManager isn't already installed, install it; otherwise skip."* This avoids re-installing every time you knit the document.

```{r install, eval=FALSE}
# Install BiocManager (the gateway to Bioconductor packages) if missing
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Use BiocManager to install metabolomicsWorkbenchR from Bioconductor
BiocManager::install("metabolomicsWorkbenchR")
```

> **Note:** `eval=FALSE` in the chunk header tells RMarkdown *not* to run this chunk when knitting. You only need to install a package once per machine; running `install.packages()` every knit would be slow and unnecessary.

## 1.2 Load the library

`library()` attaches a package's functions to your current R session. You need to do this **every time** you start a new session, even if the package is already installed.

```{r load_lib}
library(metabolomicsWorkbenchR)
```

---

# Step 2: Explore what the API can return

Before running queries, it helps to see what kinds of outputs the API supports. The package ships with a built-in data frame called `output_item` that lists every valid context/input/output combination.

```{r explore, eval=FALSE}
# Opens an interactive viewer in RStudio showing all valid query types
View(output_item)
```

> **Tip:** `View()` only works inside RStudio. If you're running this from plain R, use `head(output_item)` or `print(output_item)` instead.

---

# Step 3: Run a query

## 3.1 Define the analyte of interest

We'll search for **butyric acid**, a short-chain fatty acid produced by gut bacteria and frequently studied in microbiome-metabolomics work.

```{r analyte}
# The name we want to look up in the RefMet reference database
study_id <- "ST002681"
```

## 3.2 URL-encode the analyte name

API endpoints expect special characters (spaces, slashes, etc.) to be **percent-encoded**, so a space becomes `%20`. `utils::URLencode()` handles this for us. We set `reserved = TRUE` so it also encodes characters like `/` and `&` that have special meaning in URLs.

```{r encode}
#encode for analytes
# analyte_enc <- utils::URLencode(analyte, reserved = TRUE)
# analyte_enc  # have a look at what the encoded version looks like

#encode for study_id
# study_id_enc <- utils::URLencode(study_id, reserved = TRUE)
```

## 3.3 Query RefMet by name

`do_query()` is the workhorse function: it builds the API URL, sends the request, parses the JSON response, and hands you back a data frame.

The four arguments describe *what* you're asking for:

- `context = "refmet"`: query the RefMet reference database (which standardizes metabolite naming)
- `input_item = "name"`: we're searching by name
- `input_value = analyte_enc`: the (encoded) search term
- `output_item = "refmet_exact"`: we want an *exact/close-match* RefMet entry (vs. a fuzzy match)

```{r query}
df_study_data <- do_query(
  context     = "study",
  input_item  = "study_id",
  input_value = study_id,
  output_item = "data"
)

# Inspect the result
head(df_study_data)
```

## Save metabolomics data as CSV files

The study contains two analysis datasets: one from positive ion mode and one from negative ion mode. We will save each dataset as a separate CSV file inside an `output` folder in the current working directory.

```{r save_csv}
# Create an output folder if it does not already exist
if (!dir.exists("output")) {
  dir.create("output")
}

# Save positive ion mode data
write.csv(
  df_study_data$AN004353,
  file = "output/ST002681_AN004353_positive.csv",
  row.names = FALSE
)

# Save negative ion mode data
write.csv(
  df_study_data$AN004354,
  file = "output/ST002681_AN004354_negative.csv",
  row.names = FALSE
)
```










## Prepare baseline TMAO metabolomics data

The metabolomics dataset contains repeated measurements from the Exercise
substudy. For this analysis, we define the first metabolomics timepoint
(`E11`, or `11` if that naming convention occurs) as the baseline
metabolomics visit.

```{r load_metabolomics}

# Load metabolomics data
metab <- readr::read_csv(
  file.path(
    project_dir,
    "output",
    "Metabolomics.csv"
  ),
  show_col_types = FALSE
)

# Rename the first column, which contains sample IDs
names(metab)[1] <- "MetabSampleID"

# Check dimensions and first sample IDs
dim(metab)
head(metab$MetabSampleID)
```

## Select baseline metabolomics visit and TMAO

```{r select_baseline_tmao}

tmao_baseline <- metab %>%

  # Keep baseline metabolomics samples:
  # IDs ending in -E11 or -11
  filter(
    grepl("-(E)?11$", MetabSampleID)
  ) %>%

  # Extract participant ID from the sample ID
  mutate(
    SubjectID = sub(
      "-(E)?11$",
      "",
      MetabSampleID
    ),

    MetabVisitID = sub(
      "^.*-",
      "",
      MetabSampleID
    )
  ) %>%

  # Keep only variables needed for the TMAO analysis
  select(
    SubjectID,
    MetabSampleID,
    MetabVisitID,
    TMAO_raw = `Trimethylamine N-oxide (TMAO)`
  )

# Inspect
dim(tmao_baseline)
head(tmao_baseline)
```

## Check baseline TMAO data

```{r check_baseline_tmao}

tmao_baseline %>%
  summarise(
    n_samples = n(),
    n_subjects = n_distinct(SubjectID),
    n_missing_TMAO = sum(is.na(TMAO_raw)),
    n_zero_TMAO = sum(TMAO_raw == 0, na.rm = TRUE),
    min_TMAO = min(TMAO_raw, na.rm = TRUE),
    median_TMAO = median(TMAO_raw, na.rm = TRUE),
    max_TMAO = max(TMAO_raw, na.rm = TRUE)
  )
```

## Log-transform TMAO

TMAO abundance is right-skewed, so TMAO is log2-transformed before
statistical analyses.

```{r log_transform_tmao}

tmao_baseline <- tmao_baseline %>%
  mutate(
    TMAO_log2 = log2(TMAO_raw)
  )

# Inspect transformed values
summary(tmao_baseline$TMAO_raw)
summary(tmao_baseline$TMAO_log2)
```

## Save baseline TMAO dataset

The resulting dataset contains one row per participant and is ready to
merge with the microbiome dataset using `SubjectID`.

```{r save_baseline_tmao}

readr::write_csv(
  tmao_baseline,
  file.path(
    project_dir,
    "output",
    "iPOP_TMAO_baseline_E11.csv"
  )
)
```

