---
title: "Tutorial 1: Loading iPOP / HMP2 Abundance + Subject Metadata"
author: "Written by: Keisha Harrison | Reviewed by: OHSU CDSSR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, error = TRUE)
```

# Background material

## About iHMP CFDE

The **Integrative Human Microbiome Project (iHMP / HMP2)** Type 2 Diabetes study followed a cohort of participants longitudinally, collecting gut microbiome 16S sequencing data alongside rich clinical metadata. The data live on Stanford's Integrated Personal Omics (iPOP) data portal and are accessible as plain web tables.

Note: the same data are also distributed as **HMP2Data**, a Bioconductor Experiment Data package containing the iHMP 16S rRNA sequencing data. Its raw files come straight from the iHMP Data Analysis and Coordination Center. We download from the web tables directly here so you can see each step, but that package is a convenient alternative!

## About 16S rRNA sequencing

The **16S rRNA gene** codes for part of the ribosome and is present in all bacteria and archaea, which is what makes it useful as a universal marker. It contains **conserved regions** (slowly evolving, nearly identical across species) and **variable regions** (quickly evolving, different between species).

Sequencing this gene tells you about *composition*, not *quantity*. The number of reads assigned to a taxon depends on how deeply that sample happened to be sequenced, so it can't be read as the absolute number of cells present. Two samples sequenced to different depths aren't comparable in raw counts. Converting each sample's counts to a **relative abundance** puts every sample on the same 0-to-1 scale, which is what makes them comparable.

Those counts are tallied per **amplicon sequence variant (ASV)**: an exact, error-corrected DNA sequence recovered from the sequencing data. ASVs are then assigned to named taxa, which is the form the data arrive in for this course.

# Tutorial overview

## About Tutorial 1

In this first tutorial you will learn how to pull in the two core data components you'll need throughout the rest of the course:

- An **abundance table** (samples × taxa), downloaded as a tab-delimited text file. Each row is a sample, each column is a taxon (aka OTU/ASV), and each cell is the relative abundance of that taxon within that sample. Note that the conversion from reads to relative abundance has already been done for you.
- A **subject metadata table** (one row per participant), scraped from an HTML table. It holds stable attributes like age at enrollment, sex, race, and insulin-resistant/insulin-sensitive (IR/IS) classification.

By the end you should be able to:

- Import both datasets into R
- Verify each loaded correctly by checking its dimensions, column names, and **column types**
- Save them as an `.RData` file so the next tutorial can pick them up

**Prerequisites:** R, RStudio, internet access (the data are downloaded live from Stanford-hosted URLs), and the packages listed below.

---

# Step 1: Set up the working environment

## 1.1 Load required libraries

We rely on three packages here. `library()` attaches a package's functions to your session, and you need to install each one once with `install.packages()` before `library()` will find it.

```{r libs}
library(rvest)    # web scraping: pulls tables out of HTML pages
library(janitor)  # tidy-up helpers: standardizing messy column names
library(readr)    # fast, friendly reading of delimited text into tibbles
```

**Where each function in this tutorial comes from.** Throughout the course we call functions such as `read_tsv()`, but it is worth knowing which package each one comes from, because that is what makes a dependency problem easy to diagnose.

| Package | Functions used in this tutorial |
|---|---|
| `rvest` | `read_html()`, `html_element()`, `html_table()` |
| `readr` | `read_tsv()` |
| `janitor` | `clean_names()` |
| **base R** (nothing to load) | `c()`, `dim()`, `head()`, `str()`, `identical()`, `names()`, `stopifnot()`, `save()`, and the native pipe |

> **Pipes! The pipe is base R, not a package.** `|>` is R's *native* pipe, built into the language since R 4.1, so it works with no packages attached at all. You may also see `%>%`, the `magrittr` pipe, in other people's code; `dplyr` re-exports it, so that one *does* require a package. They behave the same way for everything we do in this course. The `|>` takes the result on its left and feeds it in as the first argument on its right, which lets you read a chain of transformations top to bottom instead of inside-out. (RStudio will type it for you with `Ctrl + Shift + M` on Windows or `Cmd + Shift + M` on Mac. If that inserts `%>%` instead, turn on *Tools -> Global Options -> Code -> Use native pipe operator*.)

> **Writing `package::function()`.** You'll see `janitor::clean_names()` below. The `::` operator says "use the `clean_names` function from the `janitor` package specifically." It's useful for two reasons: (i) it makes the source obvious to a reader, and (ii) it protects you when two attached packages export a function with the same name. We use it here and there for clarity, but since `janitor` is attached, plain `clean_names()` would work too.

---

# Step 2: Import the data

## 2.1 Load the abundance table

The abundance file is a plain tab-separated text file. `readr::read_tsv()` downloads it (since we pass a URL instead of a local path) and parses it into a tibble. Setting `show_col_types = FALSE` just suppresses the noisy startup message that lists every column's type.

> **What's a tibble?** A tibble is the tidyverse's take on a data frame, and it genuinely *is* one: run `class()` on a tibble and you'll see `data.frame` in the list, so you can use it anywhere a `data.frame` is expected. It simply behaves better. It prints a short readable preview instead of dumping all 855 rows into your console, it lists each column's type beneath its name, and it never silently converts strings to factors.
>
> **Everything we import in this course is a tibble.** Both tables in this tutorial arrive as tibbles and stay that way through Tutorial 2. There is exactly one situation where we step down to a base `data.frame`: when something needs **row names**, which tibbles deliberately don't carry. That comes up twice in Tutorial 2, both times when handing data to `phyloseq`, and we flag it in both places.

```{r abundance}
# iPOP url where the abundance data lives
abundance_url <- "http://ipop_public.gbsc.os.scg.stanford.edu/HMP/gut_16s_abundance/gut_16s_abundance.txt"

# read_tsv() comes from readr. Rows are samples, columns are taxa, and each cell
# is the relative abundance (proportion of reads) of that taxon in that sample.
# Heads up for Tutorial 2: the columns span several taxonomic ranks at once
# (phylum through genus), each prefixed with its rank, e.g. "genus_Bacteroides".
abundance <- readr::read_tsv(abundance_url, show_col_types = FALSE)

# Look at the first few rows to confirm the import worked
head(abundance)

dim(abundance)   # expect 855 samples x 97 columns (1 ID column + 96 taxa)
```

Each row is a sample (`SampleID`), and every other column is a microbial taxon with its relative abundance (proportion) for that sample.

## 2.2 Load the subject metadata

The subject metadata isn't offered as a clean download. It is embedded in an HTML page, so we fetch the page and extract the table out of it. To do so we use three `rvest` functions:

1. `rvest::read_html()` fetches the page and parses it into an HTML document
2. `rvest::html_element("table")` picks out the first `<table>` element in that document
3. `rvest::html_table()` converts that element into a tibble

We can then standardize the column names with `janitor::clean_names()`.

> Note the orientation of the **abundance** table loaded above: it arrives **samples-as-rows**, but `phyloseq` expects **taxa-as-rows**, so Tutorial 2 transposes it. Keeping track of which way a matrix is facing is one of the most common sources of confusion in microbiome analysis, so it's worth noticing now.

```{r metadata}
subject_url <- "http://hmp2-data.stanford.edu/script.php?table=subject"

# The column names we expect after cleaning, in order. We use this vector as a
# contract: if the source page ever changes, the check below fails loudly
# instead of silently shifting our data into the wrong columns.
subject_cols <- c(
  "subject_id",
  "study",
  "race",
  "sex",
  "age",
  "bmi",
  "sspg",                  # steady-state plasma glucose (insulin sensitivity marker)
  "ir_is_classification"   # "IR" (insulin-resistant), "IS" (insulin-sensitive)
)

metadata <- rvest::read_html(subject_url) |>        # rvest: download + parse the HTML page
  rvest::html_element("table") |>                   # rvest: grab the first <table> on the page
  rvest::html_table(header = TRUE, fill = TRUE) |>  # rvest: to a tibble; first row IS a header
  janitor::clean_names()                            # janitor: "IR_IS_classification" -> "ir_is_classification"

# Sanity check! Fail loudly if the scraped columns aren't what we expect
stopifnot(identical(names(metadata), subject_cols))

head(metadata)

dim(metadata)      # expect 107 subjects x 8 columns
str(metadata)      # confirm age / bmi / sspg came in as NUMERIC, not character
```

> **Why `header = TRUE`?** The source HTML table *does* carry a real header row (`SubjectID`, `Study`, `Race`, ...). If you set `header = FALSE`, two things quietly go wrong:
>
> 1. That header row is read in as if it were **a subject**, so you get 108 rows instead of 107 and one phantom participant literally named "SubjectID".
> 2. Because the first value in every column is now text, **every column is typed `character`**, including `age`, `bmi`, and `sspg`. Any downstream `mean()` or `summary()` on those would fail or return nonsense (you'd see it break in Tutorial 3).
>
> Reading the header properly gives us 107 real subjects with numeric `age` / `bmi` / `sspg`. Try flipping it to `FALSE` and rerunning `dim()` and `str()` to see both problems for yourself.

> **Why still use `janitor::clean_names()`?** The scraped names are valid R names but aren't consistent (`SubjectID`, `IR_IS_classification`). `clean_names()` converts them all to lowercase snake_case in one step, which is what the rest of the course expects. It replaces the need to rename columns by position.

**SSPG**, or steady-state plasma glucose, measures how much glucose is actively being cleared from your bloodstream into your tissues (like muscles) under the constant presence of insulin. A lower SSPG is associated with higher insulin sensitivity and vice versa for insulin resistance.

---

# Step 3: Save the loaded objects for the next tutorial

`save()` (base R) writes one or more R objects into a single `.RData` file. Tutorial 2 begins by `load()`-ing that file, so you don't have to re-download anything each time you pick the course back up.

```{r save}
save(
  abundance,
  metadata,
  file = "output/Micro_abundance.RData"
)
```

The file is now in your **current working directory**. Run `getwd()` in the console to see where that is, or in RStudio use the **Files** pane: **More -> Show Folder in New Window**. 

---

**Next:** continue to **Tutorial 2: Inspecting data and creating a phyloseq object**, where you'll explore these tables, reconcile their identifiers, and combine them into a single `phyloseq` object for downstream analysis.
