---
title: "iPOP Gut Microbiome, TMAO, and Insulin Resistance Analysis"
author: "Devendra Paudel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

# Analysis objective

The objective of this analysis is to examine associations of gut microbial
composition and circulating trimethylamine N-oxide (TMAO) with insulin
resistance status in the iPOP cohort.

The microbiome and TMAO datasets are analyzed separately to maximize the
available sample size.

For the microbiome analysis, Visit 01 is used as the first available
microbiome visit. Participants classified as insulin resistant (IR) or
insulin sensitive (IS) are included.

For the TMAO analysis, E11 is used as the first available metabolomics
visit. Participants classified as IR or IS are included.

Microbiome analyses include:

1. Alpha diversity and IR/IS status.
2. Beta diversity and IR/IS status.
3. Genus-level associations with IR/IS status using logistic regression.

TMAO is analyzed separately for its association with IR/IS status using
logistic regression.

Age, sex, and BMI are included as covariates in adjusted models.

For genus-specific analyses, genera are prevalence-filtered, zeros are
replaced with a small pseudocount, relative abundances are CLR-transformed,
and CLR values are standardized before logistic regression.

Relative abundance data are retained without CLR transformation for
alpha- and beta-diversity analyses.

---

# 1. Load required libraries

```{r libraries}

# Clear workspace
rm(list = ls())

library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(phyloseq)
library(vegan)
library(tibble)

# Main project directory
project_dir <- "/Users/paude/Documents/HMDD_project"

# Output directory
output_dir <- file.path(
  project_dir,
  "output"
)

# Figure directory
figures_dir <- file.path(
  output_dir,
  "figures"
)

# Create figure directory if it does not already exist
dir.create(
  figures_dir,
  showWarnings = FALSE,
  recursive = TRUE
)

# Prevalence threshold for genus-specific analyses
prev_cut <- 0.20
```

---

# 2. Load first-visit microbiome data

## 2.1 Load Visit 01 metadata

```{r load_microbiome_metadata}

micro_meta <- read_csv(
  file.path(
    output_dir,
    "iPOP_16S_visit01_metadata.csv"
  ),
  show_col_types = FALSE
)

# Inspect
dim(micro_meta)

head(micro_meta)

names(micro_meta)
```

---

## 2.2 Load Visit 01 genus-level phyloseq object

```{r load_microbiome_phyloseq}

ps_16s <- readRDS(
  file.path(
    output_dir,
    "iPOP_16S_visit01_genus_relative_abundance.rds"
  )
)

# Inspect
ps_16s

nsamples(ps_16s)

ntaxa(ps_16s)

# Relative abundance should sum to 1
summary(
  sample_sums(ps_16s)
)

all(
  abs(
    sample_sums(ps_16s) - 1
  ) < 1e-9
)
```

---

# 3. Prepare microbiome analysis population

Participants classified as IR or IS are retained.

Complete-case restrictions for age, sex, and BMI are applied separately
for adjusted analyses.

IR is coded as 1 and IS as 0.

```{r prepare_microbiome_population}

# Check IR/IS classification before exclusions
micro_meta %>%
  count(
    ir_is_classification
  )

# Keep participants with known IR/IS classification
micro_iris <- micro_meta %>%

  filter(
    ir_is_classification %in%
      c("IR", "IS")
  ) %>%

  mutate(

    # IR = 1, IS = 0
    IR_binary = if_else(
      ir_is_classification == "IR",
      1,
      0
    ),

    # IS reference category
    ir_is_classification = factor(
      ir_is_classification,
      levels = c(
        "IS",
        "IR"
      )
    ),

    sex = factor(
      sex
    )
  )

# Sample size
micro_iris %>%
  summarise(

    n_samples =
      n(),

    n_subjects =
      n_distinct(
        SubjectID
      ),

    n_IR =
      sum(
        IR_binary == 1
      ),

    n_IS =
      sum(
        IR_binary == 0
      )
  )

# IR/IS distribution
micro_iris %>%
  count(
    ir_is_classification
  )

# Check covariate missingness
micro_iris %>%
  summarise(

    n_total =
      n(),

    missing_age =
      sum(
        is.na(age)
      ),

    missing_sex =
      sum(
        is.na(sex)
      ),

    missing_bmi =
      sum(
        is.na(bmi)
      ),

    complete_age_sex_bmi =
      sum(
        complete.cases(
          age,
          sex,
          bmi
        )
      )
  )
```

---

# 4. Subset phyloseq object to IR/IS participants

```{r subset_phyloseq}

ps_micro_iris <- prune_samples(
  sample_names(ps_16s) %in%
    micro_iris$SampleID,
  ps_16s
)

# Remove genera completely absent after restricting participants
ps_micro_iris <- prune_taxa(
  taxa_sums(ps_micro_iris) > 0,
  ps_micro_iris
)

# Check
ps_micro_iris

nsamples(
  ps_micro_iris
)

ntaxa(
  ps_micro_iris
)

# Put metadata in same order as phyloseq samples
micro_iris <- micro_iris %>%

  filter(
    SampleID %in%
      sample_names(
        ps_micro_iris
      )
  ) %>%

  arrange(
    match(
      SampleID,
      sample_names(
        ps_micro_iris
      )
    )
  )

# Confirm alignment
stopifnot(
  all(
    micro_iris$SampleID ==
      sample_names(
        ps_micro_iris
      )
  )
)
```

---

# 5. Alpha diversity

Shannon diversity is calculated using genus-level relative abundance data.

The prevalence filter used later for genus-specific analyses is not applied
before calculation of Shannon diversity.

```{r alpha_diversity}

alpha_temp <- estimate_richness(
  ps_micro_iris,
  measures = "Shannon"
)

alpha_data <- tibble(

  SampleID =
    rownames(
      alpha_temp
    ),

  Shannon =
    alpha_temp$Shannon

) %>%

  left_join(
    micro_iris,
    by = "SampleID"
  )

# Inspect
alpha_data %>%

  select(
    SampleID,
    SubjectID,
    ir_is_classification,
    Shannon
  ) %>%

  head()
```

---

## 5.1 Prepare alpha-diversity model population

```{r alpha_model_population}

alpha_model_data <- alpha_data %>%

  filter(
    !is.na(Shannon),
    !is.na(age),
    !is.na(sex),
    !is.na(bmi)
  ) %>%

  mutate(

    # Standardize Shannon within actual model population
    Shannon_z = as.numeric(
      scale(
        Shannon
      )
    ),

    sex = factor(
      sex
    )
  )

# Sample size
alpha_model_data %>%
  summarise(

    n =
      n(),

    n_IR =
      sum(
        IR_binary == 1
      ),

    n_IS =
      sum(
        IR_binary == 0
      )
  )
```

---

## 5.2 Logistic regression for Shannon diversity

```{r alpha_logistic}

model_alpha <- glm(

  IR_binary ~
    Shannon_z +
    age +
    sex +
    bmi,

  family =
    binomial(),

  data =
    alpha_model_data,

  control =
    glm.control(
      maxit = 100
    )
)

summary(
  model_alpha
)

model_alpha$converged
```

---

## 5.3 Extract Shannon diversity result

```{r alpha_results}

alpha_coef <- summary(
  model_alpha
)$coefficients[
  "Shannon_z",
]

alpha_result <- tibble(

  variable =
    "Shannon diversity",

  n =
    nobs(
      model_alpha
    ),

  beta =
    unname(
      alpha_coef[
        "Estimate"
      ]
    ),

  SE =
    unname(
      alpha_coef[
        "Std. Error"
      ]
    ),

  OR =
    exp(
      unname(
        alpha_coef[
          "Estimate"
        ]
      )
    ),

  CI_lower =
    exp(
      unname(
        alpha_coef[
          "Estimate"
        ]
      ) -
        1.96 *
        unname(
          alpha_coef[
            "Std. Error"
          ]
        )
    ),

  CI_upper =
    exp(
      unname(
        alpha_coef[
          "Estimate"
        ]
      ) +
        1.96 *
        unname(
          alpha_coef[
            "Std. Error"
          ]
        )
    ),

  p_value =
    unname(
      alpha_coef[
        "Pr(>|z|)"
      ]
    ),

  converged =
    model_alpha$converged
)

alpha_result
```

---

## 5.4 Plot Shannon diversity

```{r alpha_final_plot}

alpha_model_label <- paste0(

  "Adjusted OR per 1-SD higher Shannon = ",

  sprintf(
    "%.2f",
    alpha_result$OR
  ),

  " (95% CI ",

  sprintf(
    "%.2f",
    alpha_result$CI_lower
  ),

  "–",

  sprintf(
    "%.2f",
    alpha_result$CI_upper
  ),

  "), P = ",

  format.pval(
    alpha_result$p_value,
    digits = 2,
    eps = 0.001
  )
)

alpha_plot <- ggplot(

  alpha_model_data,

  aes(
    x = ir_is_classification,
    y = Shannon,
    fill = ir_is_classification,
    color = ir_is_classification
  )

) +

  geom_boxplot(
    alpha = 0.35,
    outlier.shape = NA,
    width = 0.55
  ) +

  geom_jitter(
    width = 0.12,
    height = 0,
    size = 2.5,
    alpha = 0.8
  ) +

  scale_fill_manual(
    values = c(
      "IS" = "#4C78A8",
      "IR" = "#E45756"
    )
  ) +

  scale_color_manual(
    values = c(
      "IS" = "#4C78A8",
      "IR" = "#E45756"
    )
  ) +

  labs(

    x =
      "Insulin resistance classification",

    y =
      "Shannon diversity",

    title =
      "Gut microbial alpha diversity by IR/IS status",

    subtitle =
      alpha_model_label
  ) +

  theme_classic() +

  theme(
    legend.position =
      "none"
  )

alpha_plot
```

---

## 5.5 Save alpha-diversity plot

```{r save_alpha_plot}

ggsave(

  filename = file.path(
    figures_dir,
    "iPOP_Shannon_IRIS.png"
  ),

  plot =
    alpha_plot,

  width =
    7,

  height =
    5,

  units =
    "in",

  dpi =
    300,

  bg =
    "white"
)
```

---

# 6. Beta diversity

Bray-Curtis dissimilarity is calculated using genus-level relative
abundance data.

```{r prepare_beta_diversity}

# Complete-case population
beta_meta <- micro_iris %>%

  filter(
    !is.na(age),
    !is.na(sex),
    !is.na(bmi)
  ) %>%

  mutate(
    sex = factor(
      sex
    )
  )

# Sample size
beta_meta %>%
  summarise(

    n =
      n(),

    n_IR =
      sum(
        IR_binary == 1
      ),

    n_IS =
      sum(
        IR_binary == 0
      )
  )

# Extract relative-abundance matrix
otu_rel <- as(
  otu_table(
    ps_micro_iris
  ),
  "matrix"
)

# Samples must be rows
if (
  taxa_are_rows(
    ps_micro_iris
  )
) {

  otu_rel <- t(
    otu_rel
  )
}

# Keep beta-analysis participants
otu_rel <- otu_rel[
  beta_meta$SampleID,
  ,
  drop = FALSE
]

# Confirm sample order
stopifnot(
  all(
    rownames(
      otu_rel
    ) ==
      beta_meta$SampleID
  )
)

# Bray-Curtis dissimilarity
bray_dist <- vegdist(
  otu_rel,
  method = "bray"
)
```

---

## 6.1 PERMANOVA

```{r permanova}

beta_meta <- as.data.frame(
  beta_meta
)

rownames(
  beta_meta
) <- beta_meta$SampleID

# Put metadata in exact distance-matrix order
beta_meta <- beta_meta[

  attr(
    bray_dist,
    "Labels"
  ),

  ,

  drop = FALSE
]

# Confirm alignment
stopifnot(
  all(
    rownames(
      beta_meta
    ) ==
      attr(
        bray_dist,
        "Labels"
      )
  )
)

set.seed(
  1234
)

permanova_result <- adonis2(

  bray_dist ~
    ir_is_classification +
    age +
    sex +
    bmi,

  data =
    beta_meta,

  permutations =
    9999,

  by =
    "margin"
)

permanova_result
```

---

## 6.2 Extract IR/IS PERMANOVA result

```{r permanova_extract}

permanova_table <- permanova_result %>%

  as.data.frame() %>%

  rownames_to_column(
    var = "term"
  )

permanova_table

permanova_iris <- permanova_table %>%

  filter(
    term ==
      "ir_is_classification"
  )

permanova_iris
```

---

## 6.3 Check homogeneity of multivariate dispersion

```{r betadisper}

dispersion_model <- betadisper(

  bray_dist,

  group =
    beta_meta$ir_is_classification
)

set.seed(
  1234
)

dispersion_test <- permutest(

  dispersion_model,

  permutations =
    9999
)

dispersion_test
```

A non-significant dispersion test indicates that there is no evidence
that within-group multivariate dispersion differs between IR and IS.

---

## 6.4 PCoA visualization

The PCoA is an unconstrained descriptive visualization. Adjusted PERMANOVA
results are reported separately.

```{r pcoa}

pcoa_result <- cmdscale(

  bray_dist,

  k =
    2,

  eig =
    TRUE,

  add =
    TRUE
)

# Plotting dataframe
pcoa_data <- tibble(

  SampleID =
    rownames(
      pcoa_result$points
    ),

  PCoA1 =
    pcoa_result$points[
      ,
      1
    ],

  PCoA2 =
    pcoa_result$points[
      ,
      2
    ]

) %>%

  left_join(

    beta_meta %>%
      select(
        SampleID,
        ir_is_classification
      ),

    by =
      "SampleID"
  )

# Percent variation
positive_eigenvalues <-
  pcoa_result$eig[
    pcoa_result$eig > 0
  ]

PCoA1_percent <-
  100 *
  pcoa_result$eig[1] /
  sum(
    positive_eigenvalues
  )

PCoA2_percent <-
  100 *
  pcoa_result$eig[2] /
  sum(
    positive_eigenvalues
  )

# Plot
beta_plot <- ggplot(

  pcoa_data,

  aes(
    x = PCoA1,
    y = PCoA2,
    color = ir_is_classification
  )

) +

  geom_point(
    size = 3,
    alpha = 0.85
  ) +

  scale_color_manual(
    values = c(
      "IS" = "#4C78A8",
      "IR" = "#E45756"
    )
  ) +

  labs(

    x = paste0(
      "PCoA1 (",
      round(
        PCoA1_percent,
        1
      ),
      "%)"
    ),

    y = paste0(
      "PCoA2 (",
      round(
        PCoA2_percent,
        1
      ),
      "%)"
    ),

    color =
      "IR/IS classification",

    title =
      "Bray-Curtis beta diversity by IR/IS status"
  ) +

  theme_classic()

beta_plot
```

---

## 6.5 Save beta-diversity plot

```{r save_beta_plot}

ggsave(

  filename = file.path(
    figures_dir,
    "iPOP_BrayCurtis_IRIS_PCoA.png"
  ),

  plot =
    beta_plot,

  width =
    7,

  height =
    5,

  units =
    "in",

  dpi =
    300,

  bg =
    "white"
)
```

---

# 7. Prepare genus abundance matrix for CLR analysis

```{r prepare_clr}

# Complete-case genus analysis population
micro_genus <- micro_iris %>%

  filter(
    !is.na(age),
    !is.na(sex),
    !is.na(bmi)
  ) %>%

  mutate(
    sex = factor(
      sex
    )
  )

# Sample size
micro_genus %>%
  summarise(

    n =
      n(),

    n_IR =
      sum(
        IR_binary == 1
      ),

    n_IS =
      sum(
        IR_binary == 0
      )
  )

# Extract genus abundance matrix
otu_clr_input <- as(

  otu_table(
    ps_micro_iris
  ),

  "matrix"
)

# Samples must be rows
if (
  taxa_are_rows(
    ps_micro_iris
  )
) {

  otu_clr_input <- t(
    otu_clr_input
  )
}

# Restrict to complete-case model participants
otu_clr_input <- otu_clr_input[
  micro_genus$SampleID,
  ,
  drop = FALSE
]

# Check alignment
stopifnot(
  all(
    rownames(
      otu_clr_input
    ) ==
      micro_genus$SampleID
  )
)

dim(
  otu_clr_input
)

# Missing values
sum(
  is.na(
    otu_clr_input
  )
)
```

---

## 7.1 Assign genus names

```{r genus_names}

tax_df <- as.data.frame(
  tax_table(
    ps_micro_iris
  )
)

# Match taxonomy order
tax_df <- tax_df[
  colnames(
    otu_clr_input
  ),
  ,
  drop = FALSE
]

# Genus labels
genus_labels <- as.character(
  tax_df$Taxon
)

# Replace missing labels with taxon ID
missing_genus <-
  is.na(
    genus_labels
  ) |
  genus_labels == ""

genus_labels[
  missing_genus
] <- colnames(
  otu_clr_input
)[
  missing_genus
]

# Ensure uniqueness
genus_labels <- make.unique(
  genus_labels
)

# Rename matrix columns
colnames(
  otu_clr_input
) <- genus_labels

colnames(
  otu_clr_input
)
```

---

# 8. Genus prevalence filtering

Genera are retained if detected in at least 20% of the genus-analysis
population.

```{r prevalence_filter}

genus_prevalence <- colMeans(
  otu_clr_input > 0
)

prevalence_all <- tibble(

  genus =
    names(
      genus_prevalence
    ),

  prevalence =
    as.numeric(
      genus_prevalence
    )

) %>%

  arrange(
    desc(
      prevalence
    )
  )

prevalence_all
```

---

## 8.1 Apply prevalence threshold

```{r apply_prevalence_filter}

keep_genera <-
  genus_prevalence >=
  prev_cut

n_genera_before <-
  ncol(
    otu_clr_input
  )

otu_clr_filtered <-
  otu_clr_input[
    ,
    keep_genera,
    drop = FALSE
  ]

prevalence_table <- prevalence_all %>%

  filter(
    prevalence >=
      prev_cut
  )

n_genera_after <-
  ncol(
    otu_clr_filtered
  )

n_genera_before

n_genera_after

prevalence_table
```

---

# 9. Replace zeros before CLR transformation

```{r zero_replacement}

# Smallest positive abundance
min_positive <- min(

  otu_clr_filtered[
    otu_clr_filtered > 0
  ],

  na.rm =
    TRUE
)

# Pseudocount
pseudocount <-
  min_positive / 2

pseudocount

# Replace zeros
otu_positive <-
  otu_clr_filtered

otu_positive[
  otu_positive == 0
] <- pseudocount

# Confirm no zeros
sum(
  otu_positive == 0
)
```

---

# 10. CLR transformation

```{r clr_transform}

log_abundance <- log(
  otu_positive
)

clr_matrix <- sweep(

  log_abundance,

  1,

  rowMeans(
    log_abundance
  ),

  FUN =
    "-"
)

dim(
  clr_matrix
)

# CLR should average approximately zero within participant
summary(
  rowMeans(
    clr_matrix
  )
)
```

---

# 11. Standardize CLR genus abundance

```{r standardize_clr}

genus_sd <- apply(

  clr_matrix,

  2,

  sd,

  na.rm =
    TRUE
)

# Remove zero-variance genera
clr_matrix <- clr_matrix[
  ,
  genus_sd > 0,
  drop = FALSE
]

# Standardize each genus
clr_z <- scale(
  clr_matrix
)

dim(
  clr_z
)
```

---

## 11.1 Save standardized CLR genus dataset

```{r save_clr}

clr_output <- as.data.frame(
  clr_z
) %>%

  rownames_to_column(
    var =
      "SampleID"
  ) %>%

  left_join(

    micro_genus %>%

      select(
        SampleID,
        SubjectID,
        ir_is_classification,
        age,
        sex,
        bmi,
        sspg
      ),

    by =
      "SampleID"
  )

write_csv(

  clr_output,

  file.path(
    output_dir,
    "iPOP_visit01_IRIS_genus_CLR_standardized.csv"
  )
)
```

---

# 12. Genus-specific logistic regression

Each genus is fitted in a separate adjusted logistic regression model.

IR is coded as 1 and IS as 0.

The OR therefore represents the odds of IR versus IS associated with a
one-standard-deviation higher CLR abundance of that genus.

```{r genus_logistic}

genus_list <- colnames(
  clr_z
)

genus_results_list <- lapply(

  genus_list,

  function(g) {

    model_data <-
      micro_genus

    # Add standardized genus abundance
    model_data$genus_z <-
      clr_z[
        model_data$SampleID,
        g
      ]

    fit <- tryCatch(

      glm(

        IR_binary ~
          genus_z +
          age +
          sex +
          bmi,

        family =
          binomial(),

        data =
          model_data,

        control =
          glm.control(
            maxit = 200
          )
      ),

      error =
        function(e)
          NULL
    )

    # Model failure
    if (
      is.null(
        fit
      )
    ) {

      return(

        tibble(

          genus =
            g,

          n =
            NA_integer_,

          beta =
            NA_real_,

          SE =
            NA_real_,

          OR =
            NA_real_,

          CI_lower =
            NA_real_,

          CI_upper =
            NA_real_,

          p_value =
            NA_real_,

          converged =
            FALSE
        )
      )
    }

    coef_table <- summary(
      fit
    )$coefficients

    # Missing genus coefficient
    if (
      !"genus_z" %in%
      rownames(
        coef_table
      )
    ) {

      return(

        tibble(

          genus =
            g,

          n =
            nobs(
              fit
            ),

          beta =
            NA_real_,

          SE =
            NA_real_,

          OR =
            NA_real_,

          CI_lower =
            NA_real_,

          CI_upper =
            NA_real_,

          p_value =
            NA_real_,

          converged =
            fit$converged
        )
      )
    }

    beta <-
      coef_table[
        "genus_z",
        "Estimate"
      ]

    se <-
      coef_table[
        "genus_z",
        "Std. Error"
      ]

    p <-
      coef_table[
        "genus_z",
        "Pr(>|z|)"
      ]

    tibble(

      genus =
        g,

      n =
        nobs(
          fit
        ),

      beta =
        beta,

      SE =
        se,

      OR =
        exp(
          beta
        ),

      CI_lower =
        exp(
          beta -
            1.96 * se
        ),

      CI_upper =
        exp(
          beta +
            1.96 * se
        ),

      p_value =
        p,

      converged =
        fit$converged
    )
  }
)

genus_results <- bind_rows(
  genus_results_list
)
```

---

## 12.1 Add prevalence and FDR correction

```{r genus_results_fdr}

genus_results <- genus_results %>%

  left_join(
    prevalence_table,
    by = "genus"
  ) %>%

  mutate(

    p_FDR = p.adjust(
      p_value,
      method = "BH"
    )

  ) %>%

  arrange(
    p_value
  )

genus_results
```

---

## 12.2 Check logistic regression convergence

```{r genus_convergence}

genus_results %>%
  count(
    converged
  )

# Failed or unavailable models
genus_results %>%

  filter(
    !converged |
      is.na(
        p_value
      )
  )
```

---

## 12.3 Nominally associated genera

These are exploratory associations with nominal P < 0.05.

```{r nominal_genera}

genus_nominal_results <- genus_results %>%

  filter(
    !is.na(
      p_value
    ),
    p_value < 0.05
  ) %>%

  select(
    genus,
    n,
    prevalence,
    beta,
    SE,
    OR,
    CI_lower,
    CI_upper,
    p_value,
    p_FDR,
    converged
  )

genus_nominal_results
```

---

## 12.4 FDR-significant genera

```{r fdr_genera}

genus_fdr_results <- genus_results %>%

  filter(
    !is.na(
      p_FDR
    ),
    p_FDR < 0.05
  ) %>%

  select(
    genus,
    n,
    prevalence,
    beta,
    SE,
    OR,
    CI_lower,
    CI_upper,
    p_value,
    p_FDR,
    converged
  )

genus_fdr_results
```

---

## 12.5 Prepare nominal genus results for forest plot

Only genera with nominal P < 0.05 are included in this visualization.

The plot is intended to show the direction and magnitude of exploratory
genus-specific associations.

```{r genus_plot_data}

genus_plot_data <- genus_results %>%

  filter(

    converged,

    !is.na(
      p_value
    ),

    p_value < 0.05,

    !is.na(
      OR
    ),

    !is.na(
      CI_lower
    ),

    !is.na(
      CI_upper
    ),

    is.finite(
      OR
    ),

    is.finite(
      CI_lower
    ),

    is.finite(
      CI_upper
    ),

    OR > 0,

    CI_lower > 0,

    CI_upper > 0
  ) %>%

  mutate(

    direction = case_when(

      OR > 1 ~
        "Higher odds of IR",

      OR < 1 ~
        "Lower odds of IR",

      TRUE ~
        "OR = 1"
    ),

    genus_plot = reorder(
      genus,
      OR
    )
  )

genus_plot_data
```

---

## 12.6 Forest plot of nominally associated genera

The dashed line at OR = 1 indicates no association.

Red indicates an OR above 1 and therefore higher odds of IR per
one-standard-deviation higher CLR abundance.

Blue indicates an OR below 1 and therefore lower odds of IR per
one-standard-deviation higher CLR abundance.

```{r genus_forest_plot}

if (
  nrow(
    genus_plot_data
  ) > 0
) {

  genus_forest_plot <- ggplot(

    genus_plot_data,

    aes(
      x = OR,
      y = genus_plot,
      color = direction
    )

  ) +

    geom_vline(

      xintercept =
        1,

      linetype =
        "dashed",

      linewidth =
        0.6,

      color =
        "grey40"
    ) +

    geom_segment(

      aes(
        x = CI_lower,
        xend = CI_upper,
        y = genus_plot,
        yend = genus_plot
      ),

      linewidth =
        0.7
    ) +

    geom_point(
      size = 3
    ) +

    scale_x_log10() +

    scale_color_manual(

      values = c(

        "Lower odds of IR" =
          "#4C78A8",

        "Higher odds of IR" =
          "#E45756",

        "OR = 1" =
          "grey40"
      )
    ) +

    labs(

      x =
        "Adjusted OR for IR per 1-SD higher CLR abundance",

      y =
        NULL,

      color =
        "Direction",

      title =
        "Genus-specific associations with insulin resistance",

      subtitle =
        "Nominal associations (P < 0.05); adjusted for age, sex, and BMI"
    ) +

    theme_classic() +

    theme(

      legend.position =
        "bottom",

      axis.text.y =
        element_text(
          size = 10
        )
    )

  genus_forest_plot

} else {

  message(
    "No genera had nominal P < 0.05."
  )
}
```

---

## 12.7 Save genus forest plot

```{r save_genus_plot}

if (
  nrow(
    genus_plot_data
  ) > 0
) {

  genus_plot_height <- max(

    5,

    0.55 *
      nrow(
        genus_plot_data
      ) +
      2
  )

  ggsave(

    filename = file.path(
      figures_dir,
      "iPOP_genus_IRIS_nominal_forest_plot.png"
    ),

    plot =
      genus_forest_plot,

    width =
      8,

    height =
      genus_plot_height,

    units =
      "in",

    dpi =
      300,

    bg =
      "white"
  )
}
```

---

## 12.8 Save genus-specific results

```{r save_genus_results}

# All genus prevalence values
write_csv(

  prevalence_all,

  file.path(
    output_dir,
    "iPOP_genus_prevalence_visit01_IRIS.csv"
  )
)

# Retained genus prevalence values
write_csv(

  prevalence_table,

  file.path(
    output_dir,
    "iPOP_genus_prevalence_retained_visit01_IRIS.csv"
  )
)

# Full results for every tested genus
write_csv(

  genus_results,

  file.path(
    output_dir,
    "iPOP_genus_IRIS_logistic_results.csv"
  )
)

# Nominal P < 0.05 results
write_csv(

  genus_nominal_results,

  file.path(
    output_dir,
    "iPOP_genus_IRIS_nominal_p_lt_0.05.csv"
  )
)

# FDR < 0.05 results
write_csv(

  genus_fdr_results,

  file.path(
    output_dir,
    "iPOP_genus_IRIS_FDR_lt_0.05.csv"
  )
)

# Data underlying genus forest plot
write_csv(

  genus_plot_data %>%
    select(
      -genus_plot
    ),

  file.path(
    output_dir,
    "iPOP_genus_IRIS_nominal_forest_plot_data.csv"
  )
)
```

---

# 13. Load E11 TMAO data with subject metadata

```{r load_tmao}

tmao_data <- read_csv(

  file.path(
    output_dir,
    "iPOP_TMAO_E11_with_metadata.csv"
  ),

  show_col_types =
    FALSE
)

# Inspect
dim(
  tmao_data
)

head(
  tmao_data
)

# IR/IS distribution
tmao_data %>%
  count(
    ir_is_classification
  )
```

---

# 14. Prepare TMAO analysis population

Participants classified as IR or IS with non-missing TMAO, age, sex,
and BMI are retained.

```{r prepare_tmao_analysis}

tmao_iris <- tmao_data %>%

  filter(

    ir_is_classification %in%
      c(
        "IR",
        "IS"
      ),

    !is.na(
      TMAO_log2
    ),

    !is.na(
      age
    ),

    !is.na(
      sex
    ),

    !is.na(
      bmi
    )
  ) %>%

  mutate(

    # IR = 1, IS = 0
    IR_binary = if_else(
      ir_is_classification == "IR",
      1,
      0
    ),

    ir_is_classification = factor(
      ir_is_classification,
      levels = c(
        "IS",
        "IR"
      )
    ),

    sex = factor(
      sex
    )
  )

# Final TMAO sample size
tmao_iris %>%

  summarise(

    n_total =
      n(),

    n_subjects =
      n_distinct(
        SubjectID
      ),

    n_IR =
      sum(
        IR_binary == 1
      ),

    n_IS =
      sum(
        IR_binary == 0
      )
  )

tmao_iris %>%
  count(
    ir_is_classification
  )
```

---

# 15. Describe TMAO by IR/IS status

```{r tmao_descriptive}

tmao_descriptive <- tmao_iris %>%

  group_by(
    ir_is_classification
  ) %>%

  summarise(

    n =
      n(),

    mean_log2_TMAO =
      mean(
        TMAO_log2,
        na.rm = TRUE
      ),

    sd_log2_TMAO =
      sd(
        TMAO_log2,
        na.rm = TRUE
      ),

    median_log2_TMAO =
      median(
        TMAO_log2,
        na.rm = TRUE
      ),

    IQR_log2_TMAO =
      IQR(
        TMAO_log2,
        na.rm = TRUE
      ),

    .groups =
      "drop"
  )

tmao_descriptive
```

---

# 16. Descriptive TMAO visualization

This figure displays the observed TMAO distribution and is not an adjusted
model-based figure.

```{r tmao_descriptive_plot}

tmao_plot <- ggplot(

  tmao_iris,

  aes(
    x = ir_is_classification,
    y = TMAO_log2,
    fill = ir_is_classification,
    color = ir_is_classification
  )

) +

  geom_boxplot(

    alpha =
      0.35,

    outlier.shape =
      NA,

    width =
      0.55
  ) +

  geom_jitter(

    width =
      0.12,

    height =
      0,

    size =
      2.5,

    alpha =
      0.8
  ) +

  scale_fill_manual(

    values = c(
      "IS" = "#4C78A8",
      "IR" = "#E45756"
    )
  ) +

  scale_color_manual(

    values = c(
      "IS" = "#4C78A8",
      "IR" = "#E45756"
    )
  ) +

  labs(

    x =
      "Insulin resistance classification",

    y =
      "Log2 TMAO",

    title =
      "Observed circulating TMAO by IR/IS status"
  ) +

  theme_classic() +

  theme(
    legend.position =
      "none"
  )

tmao_plot
```

---

## 16.1 Save descriptive TMAO plot

```{r save_tmao_descriptive_plot}

ggsave(

  filename = file.path(
    figures_dir,
    "iPOP_TMAO_IRIS_descriptive.png"
  ),

  plot =
    tmao_plot,

  width =
    7,

  height =
    5,

  units =
    "in",

  dpi =
    300,

  bg =
    "white"
)
```

---

# 17. TMAO and IR/IS logistic regression

The association between circulating TMAO and insulin resistance status is
evaluated using logistic regression adjusted for age, sex, and BMI.

Because TMAO is log2-transformed, a one-unit increase in TMAO_log2
corresponds to a two-fold higher TMAO value.

```{r tmao_logistic}

model_tmao <- glm(

  IR_binary ~
    TMAO_log2 +
    age +
    sex +
    bmi,

  family =
    binomial(),

  data =
    tmao_iris,

  control =
    glm.control(
      maxit = 100
    )
)

summary(
  model_tmao
)

model_tmao$converged
```

---

## 17.1 Extract TMAO association result

```{r tmao_result}

tmao_coef <- summary(
  model_tmao
)$coefficients[
  "TMAO_log2",
]

tmao_result <- tibble(

  predictor =
    "TMAO_log2",

  n =
    nobs(
      model_tmao
    ),

  beta =
    unname(
      tmao_coef[
        "Estimate"
      ]
    ),

  SE =
    unname(
      tmao_coef[
        "Std. Error"
      ]
    ),

  OR =
    exp(
      unname(
        tmao_coef[
          "Estimate"
        ]
      )
    ),

  CI_lower =
    exp(
      unname(
        tmao_coef[
          "Estimate"
        ]
      ) -
        1.96 *
        unname(
          tmao_coef[
            "Std. Error"
          ]
        )
    ),

  CI_upper =
    exp(
      unname(
        tmao_coef[
          "Estimate"
        ]
      ) +
        1.96 *
        unname(
          tmao_coef[
            "Std. Error"
          ]
        )
    ),

  p_value =
    unname(
      tmao_coef[
        "Pr(>|z|)"
      ]
    ),

  converged =
    model_tmao$converged
)

tmao_result
```

---

## 17.2 Calculate adjusted predicted probability of IR across TMAO

Marginal standardized predictions are calculated across the observed TMAO
range.

At each TMAO value, predictions are generated using the observed age, sex,
and BMI values of the analysis participants and then averaged.

```{r tmao_predictions}

# Sequence across observed TMAO range
tmao_grid <- seq(

  min(
    tmao_iris$TMAO_log2,
    na.rm = TRUE
  ),

  max(
    tmao_iris$TMAO_log2,
    na.rm = TRUE
  ),

  length.out =
    100
)

# Model coefficients
beta_hat <- coef(
  model_tmao
)

# Variance-covariance matrix
vcov_hat <- vcov(
  model_tmao
)

# Predictor terms without response
model_terms <- delete.response(
  terms(
    model_tmao
  )
)

# Predictions across TMAO values
tmao_prediction_list <- lapply(

  tmao_grid,

  function(x) {

    newdata <-
      tmao_iris

    # Give each participant the same TMAO value
    newdata$TMAO_log2 <-
      x

    # Model matrix
    X <- model.matrix(
      model_terms,
      data = newdata
    )

    # Align coefficient order
    X <- X[
      ,
      names(
        beta_hat
      ),
      drop = FALSE
    ]

    # Linear predictor
    eta <- as.vector(
      X %*%
        beta_hat
    )

    # Participant-specific predicted probabilities
    prob <- plogis(
      eta
    )

    # Average adjusted probability
    adjusted_probability <- mean(
      prob
    )

    # Delta-method gradient
    gradient_matrix <- sweep(

      X,

      1,

      prob *
        (
          1 -
            prob
        ),

      FUN =
        "*"
    )

    gradient <- colMeans(
      gradient_matrix
    )

    variance <- as.numeric(

      t(
        gradient
      ) %*%

        vcov_hat %*%

        gradient
    )

    SE <- sqrt(
      variance
    )

    tibble(

      TMAO_log2 =
        x,

      predicted_probability_IR =
        adjusted_probability,

      SE =
        SE,

      CI_lower =
        pmax(
          0,
          adjusted_probability -
            1.96 * SE
        ),

      CI_upper =
        pmin(
          1,
          adjusted_probability +
            1.96 * SE
        )
    )
  }
)

tmao_predictions <- bind_rows(
  tmao_prediction_list
)

tmao_predictions
```

---

## 17.3 Plot adjusted TMAO association

This figure is generated from the adjusted logistic regression model.

```{r tmao_adjusted_plot}

tmao_model_label <- paste0(

  "Adjusted OR per 2-fold higher TMAO value = ",

  sprintf(
    "%.2f",
    tmao_result$OR
  ),

  " (95% CI ",

  sprintf(
    "%.2f",
    tmao_result$CI_lower
  ),

  "–",

  sprintf(
    "%.2f",
    tmao_result$CI_upper
  ),

  "), P = ",

  format.pval(
    tmao_result$p_value,
    digits = 2,
    eps = 0.001
  )
)

tmao_adjusted_plot <- ggplot(

  tmao_predictions,

  aes(
    x = TMAO_log2,
    y = predicted_probability_IR
  )

) +

  geom_ribbon(

    aes(
      ymin = CI_lower,
      ymax = CI_upper
    ),

    alpha =
      0.20,

    fill =
      "#E45756"
  ) +

  geom_line(

    linewidth =
      1.2,

    color =
      "#E45756"
  ) +

  geom_rug(

    data =
      tmao_iris,

    aes(
      x = TMAO_log2
    ),

    inherit.aes =
      FALSE,

    sides =
      "b",

    alpha =
      0.45
  ) +

  labs(

    x =
      "Log2 TMAO",

    y =
      "Adjusted predicted probability of IR",

    title =
      "Adjusted association between circulating TMAO and insulin resistance",

    subtitle =
      tmao_model_label
  ) +

  theme_classic()

tmao_adjusted_plot
```

---

## 17.4 Save adjusted TMAO plot

```{r save_tmao_adjusted_plot}

ggsave(

  filename = file.path(
    figures_dir,
    "iPOP_TMAO_IRIS_adjusted_probability.png"
  ),

  plot =
    tmao_adjusted_plot,

  width =
    7,

  height =
    5,

  units =
    "in",

  dpi =
    300,

  bg =
    "white"
)
```

---

# 18. Save TMAO analysis data and results

```{r save_tmao_results}

# Analysis population
write_csv(

  tmao_iris,

  file.path(
    output_dir,
    "iPOP_TMAO_E11_IRIS_analysis_data.csv"
  )
)

# Descriptive statistics
write_csv(

  tmao_descriptive,

  file.path(
    output_dir,
    "iPOP_TMAO_E11_IRIS_descriptive.csv"
  )
)

# Logistic regression result
write_csv(

  tmao_result,

  file.path(
    output_dir,
    "iPOP_TMAO_E11_IRIS_logistic_result.csv"
  )
)

# Adjusted prediction data
write_csv(

  tmao_predictions,

  file.path(
    output_dir,
    "iPOP_TMAO_E11_IRIS_adjusted_predictions.csv"
  )
)
```

---

# 19. Save alpha-diversity results

```{r save_alpha_result}

# Logistic regression result
write_csv(

  alpha_result,

  file.path(
    output_dir,
    "iPOP_Shannon_IRIS_logistic_result.csv"
  )
)

# Analysis data
write_csv(

  alpha_model_data,

  file.path(
    output_dir,
    "iPOP_Shannon_IRIS_analysis_data.csv"
  )
)
```

---

# 20. Save beta-diversity results

```{r save_beta_result}

# Complete PERMANOVA table
permanova_table <-
  permanova_result %>%

  as.data.frame() %>%

  rownames_to_column(
    var = "term"
  )

write_csv(

  permanova_table,

  file.path(
    output_dir,
    "iPOP_BrayCurtis_IRIS_PERMANOVA.csv"
  )
)

# IR/IS-specific PERMANOVA result
write_csv(

  permanova_iris,

  file.path(
    output_dir,
    "iPOP_BrayCurtis_IRIS_PERMANOVA_IRIS_result.csv"
  )
)

# Dispersion result
dispersion_table <- tibble(

  statistic =
    unname(
      dispersion_test$tab[
        1,
        "F"
      ]
    ),

  p_value =
    unname(
      dispersion_test$tab[
        1,
        "Pr(>F)"
      ]
    )
)

write_csv(

  dispersion_table,

  file.path(
    output_dir,
    "iPOP_BrayCurtis_IRIS_dispersion_test.csv"
  )
)

# PCoA coordinates
write_csv(

  pcoa_data,

  file.path(
    output_dir,
    "iPOP_BrayCurtis_IRIS_PCoA_coordinates.csv"
  )
)
```

---

# 21. Save microbiome analysis metadata

```{r save_microbiome_analysis_metadata}

# All Visit 01 known IR/IS participants
write_csv(

  micro_iris,

  file.path(
    output_dir,
    "iPOP_visit01_IRIS_microbiome_analysis_metadata.csv"
  )
)

# Genus-analysis complete-case population
write_csv(

  micro_genus,

  file.path(
    output_dir,
    "iPOP_visit01_IRIS_genus_analysis_metadata.csv"
  )
)

# Beta-diversity complete-case population
write_csv(

  beta_meta,

  file.path(
    output_dir,
    "iPOP_visit01_IRIS_beta_analysis_metadata.csv"
  )
)
```

---

# 22. Analysis summary

The microbiome analysis uses first-visit (Visit 01) genus-level microbiome
data from participants classified as insulin resistant or insulin sensitive.

Shannon diversity is used to evaluate alpha diversity. Logistic regression
evaluates the association between a one-standard-deviation higher Shannon
diversity and odds of IR after adjustment for age, sex, and BMI.

Bray-Curtis dissimilarity is used to evaluate beta diversity. Overall
microbial community composition is evaluated using PERMANOVA adjusted for
age, sex, and BMI. Homogeneity of multivariate dispersion is evaluated
separately. PCoA provides an unconstrained descriptive visualization of
Bray-Curtis community structure, while adjusted PERMANOVA statistics are
reported separately.

For genus-specific analyses, genera are required to be present in at least
20% of participants in the complete-case genus-analysis population. Zeros
are replaced with one-half of the smallest observed non-zero relative
abundance among retained genera.

The filtered abundance matrix is CLR-transformed and each retained genus is
standardized to mean zero and standard deviation one.

Separate logistic regression models are fitted for each genus, with IR/IS
status as the binary outcome and age, sex, and BMI as covariates.
Benjamini-Hochberg false discovery rate correction is applied across
genus-specific tests.

All genus-specific regression results are retained in the complete results
table. For exploratory visualization of direction and magnitude, the forest
plot displays only nominally associated genera with P < 0.05. These
nominal associations should be interpreted separately from FDR-corrected
results.

The TMAO analysis is conducted separately using E11 TMAO samples with
subject metadata already attached.

Log2-transformed TMAO is evaluated for its association with IR/IS using
logistic regression adjusted for age, sex, and BMI.

The observed TMAO boxplot is retained as a descriptive visualization.
A separate model-based figure displays the marginal adjusted predicted
probability of IR across the observed TMAO range after accounting for age,
sex, and BMI.

A later targeted analysis can evaluate associations between circulating
TMAO and literature-defined TMA/TMAO-related genera among participants with
both microbiome and TMAO measurements.