---
title: "iPOP TMAO, Gut Microbiome, and Type 2 Diabetes Analysis"
author: "Devendra Paudel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

# Project objective

The overall objective of this analysis is to examine whether circulating
trimethylamine N-oxide (TMAO), a gut microbiota-related metabolite, is
associated with gut microbial composition and type 2 diabetes (T2D) status
in the iPOP cohort.

For this analysis, we will select a single timepoint with both microbiome
and metabolomics data available for the same participants.

The planned workflow is:

1. Load the processed iPOP 16S microbiome data.
2. Examine available subjects and visits.
3. Identify a common microbiome and metabolomics timepoint.
4. Obtain the matching iPOP metabolomics dataset.
5. Extract TMAO.
6. Obtain T2D/clinical phenotype information.
7. Match participants across microbiome, metabolomics, and clinical data.
8. Conduct microbiome-TMAO-T2D analyses.

---

# Step 1: Load required libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(phyloseq)

# Define the main project directory
project_dir <- "/Users/paude/Documents/HMDD_project"
```

---

# Step 2: Load processed 16S data

## 2.1 Load sample metadata

```{r load_microbiome_metadata}
micro_meta <- read_csv(
  file.path(
    project_dir,
    "output",
    "iPOP_16S_sample_metadata.csv"
  ),
  show_col_types = FALSE
)

# Inspect the data
dim(micro_meta)
head(micro_meta)
colnames(micro_meta)
```

## 2.2 Load genus-level relative abundance data

```{r load_microbiome}
ps_16s <- readRDS(
  file.path(
    project_dir,
    "output",
    "iPOP_16S_genus_relative_abundance.rds"
  )
)

# Inspect phyloseq object
ps_16s

nsamples(ps_16s)
ntaxa(ps_16s)
```

---

# Step 3: Examine microbiome sample identifiers and visits

The microbiome SampleID has already been separated into SubjectID and VisitID.
This information will later be used to match microbiome samples to the
metabolomics data.

```{r inspect_ids}
micro_meta %>%
  select(
    SampleID,
    SubjectID,
    VisitID,
    VisitNumber
  ) %>%
  head(20)
```

Check the total number of microbiome samples and participants:

```{r microbiome_sample_size}
micro_meta %>%
  summarise(
    n_samples = n(),
    n_subjects = n_distinct(SubjectID),
    n_visits = n_distinct(VisitID)
  )
```

---

# Step 4: Examine available microbiome visits

Count the number of microbiome samples available at each visit.

```{r visit_counts}
visit_counts <- micro_meta %>%
  count(
    VisitID,
    sort = TRUE,
    name = "n_samples"
  )

visit_counts
```

Also examine the numeric visit number where available.

```{r visit_number_counts}
micro_meta %>%
  filter(!is.na(VisitNumber)) %>%
  count(
    VisitNumber,
    sort = TRUE,
    name = "n_samples"
  )
```

---

# Step 5: Examine available subject-level variables

```{r inspect_subject_variables}
micro_meta %>%
  select(
    SubjectID,
    study,
    race,
    sex,
    age,
    bmi,
    sspg,
    ir_is_classification
  ) %>%
  distinct(SubjectID, .keep_all = TRUE) %>%
  head(20)
```

Check the number of subjects by insulin resistance classification:

```{r ir_is_counts}
micro_meta %>%
  distinct(SubjectID, .keep_all = TRUE) %>%
  count(
    ir_is_classification,
    name = "n_subjects"
  )
```

---

# Step 6: Check whether a T2D variable is currently available

The currently downloaded 16S metadata contains insulin resistance information,
but an explicit T2D diagnosis variable has not yet been identified.

```{r check_t2d_variables}
grep(
  pattern = "t2d|diabet|glucose|hba1c",
  x = names(micro_meta),
  value = TRUE,
  ignore.case = TRUE
)
```

If this returns:

```text
character(0)
```

then the current microbiome metadata does not contain an explicit T2D,
diabetes, glucose, or HbA1c variable.

---

# Step 7: Check participants available at Visit 01

Visit 01 is a possible candidate for the single-timepoint analysis.
We will not finalize this choice until we inspect the matching iPOP
metabolomics data.

```{r visit01}
micro_visit01 <- micro_meta %>%
  filter(VisitID == "01")

dim(micro_visit01)

n_distinct(micro_visit01$SubjectID)

micro_visit01 %>%
  select(
    SampleID,
    SubjectID,
    VisitID,
    age,
    sex,
    bmi,
    sspg,
    ir_is_classification
  ) %>%
  head(20)
```

---

# Step 8: Load baseline TMAO data and match with 16S microbiome

For the metabolomics data, `E11` is treated as the first available
metabolomics visit. For the microbiome data, Visit `01` is treated as the
first available microbiome visit.

The two datasets will be matched using the participant-level `SubjectID`.

## 8.1 Load baseline TMAO data

```{r load_baseline_tmao}

tmao_baseline <- read_csv(
  file.path(
    project_dir,
    "output",
    "iPOP_TMAO_baseline_E11.csv"
  ),
  show_col_types = FALSE
)

# Inspect
dim(tmao_baseline)
head(tmao_baseline)

# Confirm one row per participant
tmao_baseline %>%
  summarise(
    n_samples = n(),
    n_subjects = n_distinct(SubjectID)
  )
```

## 8.2 Attach full subject metadata to E11 TMAO data

The full iPOP subject metadata are joined to the E11 TMAO dataset using
SubjectID. This allows the TMAO dataset to be analyzed independently of
microbiome availability.

```{r attach_metadata_to_tmao}

# Load the full iPOP subject-level metadata
subject_env <- new.env()

load(
  file.path(
    project_dir,
    "output",
    "Micro_abundance.RData"
  ),
  envir = subject_env
)

subject_meta <- subject_env$metadata %>%
  rename(
    SubjectID = subject_id
  )

# Check full subject metadata
dim(subject_meta)
head(subject_meta)

subject_meta %>%
  count(
    ir_is_classification
  )
```

The full metadata contain subject-level characteristics including age, sex,
race, BMI, SSPG, and IR/IS classification.

```{r join_tmao_metadata}

tmao_E11_metadata <- tmao_baseline %>%

  left_join(
    subject_meta %>%
      select(
        SubjectID,
        study,
        race,
        sex,
        age,
        bmi,
        sspg,
        ir_is_classification
      ),
    by = "SubjectID"
  )

# Inspect
dim(tmao_E11_metadata)

tmao_E11_metadata %>%
  select(
    SubjectID,
    MetabSampleID,
    MetabVisitID,
    TMAO_raw,
    TMAO_log2,
    age,
    sex,
    bmi,
    sspg,
    ir_is_classification
  ) %>%
  head(20)
```

## 8.3 Check TMAO metadata completeness

```{r check_tmao_metadata}

tmao_E11_metadata %>%
  summarise(
    n_samples = n(),
    n_subjects = n_distinct(SubjectID),

    n_missing_TMAO =
      sum(is.na(TMAO_log2)),

    n_missing_age =
      sum(is.na(age)),

    n_missing_sex =
      sum(is.na(sex)),

    n_missing_bmi =
      sum(is.na(bmi)),

    n_missing_sspg =
      sum(is.na(sspg)),

    n_missing_IRIS =
      sum(is.na(ir_is_classification))
  )

# See IR / IS / Unknown distribution
tmao_E11_metadata %>%
  count(
    ir_is_classification
  )
```

## 8.4 Save E11 TMAO dataset with metadata

Keep all E11 TMAO participants at this stage, including participants with
Unknown IR/IS status. Analysis-specific exclusions will be made later.

```{r save_tmao_with_metadata}

write_csv(
  tmao_E11_metadata,
  file.path(
    project_dir,
    "output",
    "iPOP_TMAO_E11_with_metadata.csv"
  )
)
```

## 8.5 Check overlap between microbiome Visit 01 and metabolomics E11

```{r check_microbiome_metabolomics_overlap}

micro_visit01_subjects <- micro_visit01 %>%
  distinct(SubjectID) %>%
  pull(SubjectID)

tmao_E11_subjects <- tmao_baseline %>%
  distinct(SubjectID) %>%
  pull(SubjectID)

# Number of participants in each dataset
length(micro_visit01_subjects)
length(tmao_E11_subjects)

# Number of participants present in both datasets
length(
  intersect(
    micro_visit01_subjects,
    tmao_E11_subjects
  )
)

# IDs present in both datasets
intersect(
  micro_visit01_subjects,
  tmao_E11_subjects
)
```

## 8.6 Merge Visit 01 microbiome metadata with baseline TMAO

```{r merge_microbiome_tmao}

micro_tmao_matched <- micro_visit01 %>%

  # Rename microbiome visit variable for clarity
  rename(
    MicrobiomeSampleID = SampleID,
    MicrobiomeVisitID = VisitID
  ) %>%

  # Keep participants present in both datasets
  inner_join(
    tmao_baseline,
    by = "SubjectID"
  )

# Inspect final matched sample size
micro_tmao_matched %>%
  summarise(
    n_samples = n(),
    n_subjects = n_distinct(SubjectID)
  )

# Inspect selected variables
micro_tmao_matched %>%
  select(
    SubjectID,
    MicrobiomeSampleID,
    MicrobiomeVisitID,
    MetabSampleID,
    MetabVisitID,
    TMAO_raw,
    TMAO_log2,
    age,
    sex,
    race,
    bmi,
    sspg,
    ir_is_classification
  ) %>%
  head(20)
```

## 8.7 Check participants that did not match

```{r check_unmatched_subjects}

# Visit 01 microbiome participants without E11 metabolomics
micro_without_tmao <- micro_visit01 %>%
  anti_join(
    tmao_baseline,
    by = "SubjectID"
  )

# E11 metabolomics participants without Visit 01 microbiome
tmao_without_micro <- tmao_baseline %>%
  anti_join(
    micro_visit01,
    by = "SubjectID"
  )

nrow(micro_without_tmao)
nrow(tmao_without_micro)

micro_without_tmao %>%
  select(SubjectID, SampleID) %>%
  head(20)

tmao_without_micro %>%
  select(SubjectID, MetabSampleID) %>%
  head(20)
```


## 8.8 Remove participants without IR/IS classification

Participants with unknown or missing insulin resistance classification are
excluded from the final analytic dataset.

```{r filter_ir_is}

# Check classification before filtering
micro_tmao_matched %>%
  count(ir_is_classification)

# Keep only participants classified as IR or IS
micro_tmao_final <- micro_tmao_matched %>%
  filter(
    ir_is_classification %in% c("IR", "IS")
  )

# Check final sample size
micro_tmao_final %>%
  summarise(
    n_samples = n(),
    n_subjects = n_distinct(SubjectID)
  )

# Check IR/IS distribution after filtering
micro_tmao_final %>%
  count(
    ir_is_classification,
    name = "n_subjects"
  )
```

## 8.9 Subset the phyloseq object to the final matched participants

```{r subset_phyloseq_final}

matched_microbiome_ids <- micro_tmao_final$MicrobiomeSampleID

ps_micro_tmao_final <- prune_samples(
  sample_names(ps_16s) %in% matched_microbiome_ids,
  ps_16s
)

# Check final phyloseq object
ps_micro_tmao_final

nsamples(ps_micro_tmao_final)
ntaxa(ps_micro_tmao_final)

# Confirm that all final microbiome IDs are present
length(
  intersect(
    matched_microbiome_ids,
    sample_names(ps_micro_tmao_final)
  )
)

# Should return character(0)
setdiff(
  matched_microbiome_ids,
  sample_names(ps_micro_tmao_final)
)
```


## 9 Save final matched microbiome-TMAO dataset

```{r save_final_microbiome_tmao}

# Save final participant-level metadata
write_csv(
  micro_tmao_final,
  file.path(
    project_dir,
    "output",
    "iPOP_16S_visit01_TMAO_E11_IRIS_final_metadata.csv"
  )
)

# Save corresponding genus-level phyloseq object
saveRDS(
  ps_micro_tmao_final,
  file.path(
    project_dir,
    "output",
    "iPOP_16S_visit01_TMAO_E11_IRIS_final_phyloseq.rds"
  )
)
```


